---
title: "Flipper Length and Body Mass in Penguins"
author: "Rajat Vig"
date: "2026-09-23"
format:
  html:
    toc: true
---

## Introduction

The Palmer Penguins dataset contains measurements collected from penguins
observed near Palmer Station in Antarctica. In this analysis, I examine the
relationship between flipper length and body mass across the three penguin
species.

The goal is to summarize the measurements by species and visualize the
relationship between flipper length and body mass.

**Data source:** [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
created by Allison Horst using data from Palmer Station Antarctica LTER.

## Load and inspect the data

The dataset is provided by the `palmerpenguins` R package, so the analysis
does not require downloading a data file from the internet at render time.

```{r}
#| label: tbl-data-preview
#| message: false
#| warning: false

library(palmerpenguins)
library(dplyr)
library(ggplot2)

penguins_data <- penguins

cat("Rows:", nrow(penguins_data), "\n")
cat("Columns:", ncol(penguins_data), "\n")

head(penguins_data)
```

The dataset contains 344 observations and eight variables.

## Prepare the data

For this analysis, I need observations with both flipper length and body
mass available. I remove observations missing either measurement.

```{r}
#| label: tbl-clean-data

analysis_data <- penguins_data %>%
  filter(
    !is.na(flipper_length_mm),
    !is.na(body_mass_g),
    !is.na(species)
  )

cat("Rows before cleaning:", nrow(penguins_data), "\n")
cat("Rows after cleaning:", nrow(analysis_data), "\n")
cat(
  "Rows removed:",
  nrow(penguins_data) - nrow(analysis_data),
  "\n"
)

head(analysis_data)
```

The cleaned dataset contains the observations needed to examine the
relationship between flipper length and body mass.

## Summarize by species

I calculate the number of observations, mean flipper length, mean body mass,
and the correlation between the two measurements for each species.

```{r}
#| label: tbl-species-summary

species_summary <- analysis_data %>%
  group_by(species) %>%
  summarise(
    observations = n(),
    mean_flipper_mm = mean(flipper_length_mm),
    mean_body_mass_g = mean(body_mass_g),
    correlation = cor(flipper_length_mm, body_mass_g),
    .groups = "drop"
  ) %>%
  mutate(
    across(
      c(mean_flipper_mm, mean_body_mass_g, correlation),
      ~ round(.x, 2)
    )
  )

species_summary
```

The summary provides information about typical flipper length, typical body
mass, and the strength of the linear association between the two measurements
within each species.

## Visualize the relationship

The following figure plots body mass against flipper length. Each point
represents one penguin, and the fitted lines show the linear relationship
within each species.

```{r}
#| label: fig-flipper-body-mass
#| fig-cap: "Relationship between flipper length and body mass by penguin species."

ggplot(
  analysis_data,
  aes(
    x = flipper_length_mm,
    y = body_mass_g,
    color = species
  )
) +
  geom_point(alpha = 0.7) +
  geom_smooth(
    method = "lm",
    se = FALSE
  ) +
  labs(
    title = "Flipper Length and Body Mass",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  ) +
  theme_minimal()
```

## Findings

The summary statistics and figure show a positive relationship between
flipper length and body mass in the Palmer Penguins data. Penguins with
longer flippers generally have greater body mass.

The relationship can also be examined separately by species. The species
summary provides a correlation for each group, while the fitted lines in the
figure show the direction of the relationship visually.

These results are descriptive and summarize the observations in this
dataset. They do not establish that flipper length causes differences in
body mass.

## Conclusion

Using R, I loaded the Palmer Penguins dataset, inspected its structure,
removed observations with missing values for the variables of interest,
calculated species-level summaries and correlations, and created a
visualization of the relationship between flipper length and body mass.

The analysis demonstrates a reproducible R workflow because the data
processing, calculations, and figure are generated from code when the
Quarto document is rendered.
