---
title: "R and Python Analysis of Penguin Body Mass"
author: "Rajat Vig"
date: "2026-09-23"
format:
  html:
    toc: true
engine: knitr
---

## Introduction

The Palmer Penguins dataset contains measurements collected from penguins observed near Palmer Station in Antarctica. In this analysis, I use both R and Python in the same Quarto document.

The R code prepares the data and calculates species-level summaries. The Python code receives data from R, performs an additional calculation, and returns the result to R.

**Data source:** [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
collected by Dr. Kristen Gorman at Palmer Station Antarctica LTER. Licence: CC0 1.0.
Citation: Horst AM, Hill AP, Gorman KB (2020). *palmerpenguins: Palmer Archipelago
(Antarctica) penguin data.* [doi:10.5281/zenodo.3960218](https://doi.org/10.5281/zenodo.3960218).

The dataset is distributed with the `palmerpenguins` package, so the analysis does not need to download a data file from the internet during rendering.

## Load and prepare the data in R

First, I load the Palmer Penguins data in R and remove observations that are missing body mass or species.

```{r}
#| label: r-data
#| message: false
#| warning: false

library(palmerpenguins)
library(dplyr)
library(reticulate)

analysis_data <- penguins %>%
  filter(
    !is.na(species),
    !is.na(body_mass_g)
  )

cat("Rows after cleaning:", nrow(analysis_data), "\n")

head(analysis_data)
```

The cleaned dataset contains the observations needed for the analysis.

## Calculate species means in R

I calculate the mean body mass for each penguin species.

```{r}
#| label: r-summary

r_summary <- analysis_data %>%
  group_by(species) %>%
  summarise(
    observations = n(),
    mean_body_mass_g = mean(body_mass_g),
    .groups = "drop"
  ) %>%
  mutate(
    mean_body_mass_g = round(mean_body_mass_g, 1)
  )

r_summary
```

The R calculation produces a species-level summary that will be passed to Python.

## Pass the R result to Python

The `reticulate` package allows R and Python to communicate within the same Quarto document.

I pass the cleaned R data to Python as a pandas DataFrame. Python then calculates the overall mean body mass.

```{python}
#| label: python-summary

import pandas as pd

python_data = r.analysis_data

overall_mean = float(python_data["body_mass_g"].mean())

print("Python received", len(python_data), "observations")
print(f"Overall mean body mass: {overall_mean:.1f} g")
```

The Python code receives the object created in R and calculates a new result from it.

## Pass the Python result back to R

The Python result can also be accessed from R.

```{r}
#| label: r-python-result

python_mean <- py$overall_mean

cat(
  "Overall mean body mass calculated by Python:",
  round(python_mean, 1),
  "g\n"
)
```

This demonstrates the two-way workflow: R creates the data object, Python uses that object to calculate a result, and R reads the Python result.

## Visualize the data in Python

Finally, I use the Python DataFrame to create a figure showing body mass by species.

```{python}
#| label: fig-r-python-penguins
#| fig-cap: "Penguin body mass by species, using data prepared in R and visualized in Python."

import matplotlib.pyplot as plt

fig, ax = plt.subplots(figsize=(8, 5))

python_data.boxplot(
    column="body_mass_g",
    by="species",
    ax=ax
)

ax.axhline(
    overall_mean,
    color="black",
    linestyle="--",
    linewidth=1.5,
    label=f"Overall mean: {overall_mean:.0f} g"
)

ax.set_title("Penguin Body Mass by Species")
ax.set_xlabel("Species")
ax.set_ylabel("Body mass (g)")
ax.legend()

plt.suptitle("")
plt.tight_layout()
plt.show()
```

## Findings

The analysis demonstrates how R and Python can be used together in one Quarto document.

R was used to load and clean the Palmer Penguins data and calculate species-level summaries. The cleaned R data were then passed to Python, where the overall mean body mass was calculated.

The Python result was subsequently accessed from R, demonstrating that objects can be passed between the two languages through `reticulate`.

The figure shows differences in body mass among the three penguin species, along with the overall mean body mass across the cleaned observations.

These are descriptive results from the Palmer Penguins dataset and do not establish causal relationships.

## Conclusion

This post demonstrates a reproducible mixed-language workflow using R, Python, Quarto, and `reticulate`. The same computational document contains R and Python code, passes an object from R to Python, performs a calculation in Python, and passes the resulting value back to R.

Because all calculations and the figure are generated during rendering, the analysis can be reproduced from the project environment rather than relying on manually created outputs.
